Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-10 LICS21/SPCF.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

The one-step reduction is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
\begin{equation*}
\red(M,s) := \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N\\
        & \qquad \text{ and } R \neq \tsample \\
        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
        (M,s) & \text{if } M \text{ is a value}
    \end{array} \right .
\end{equation*}
\lo{Both $\to$ and $\red$ are called reduction relation.}

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
This definition of almost sure termination is equivalent to that given in \citep{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning. 
The $\tscore(r)$ construct is not relevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction stragegy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

>>>>

The one-step reduction is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
\begin{equation*}
\red(M,s) := \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N\\
        & \qquad \text{ and } R \neq \tsample \\
        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
        (M,s) & \text{if } M \text{ is a value}
    \end{array} \right .
\end{equation*}
%\lo{Both $\to$ and $\red$ are called reduction relation.}

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
This definition of almost sure termination is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning. 
The $\tscore(r)$ construct is not relevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction stragegy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

### 2021-01-10 LICS21/antitone-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.

@@

There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
%Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
%cannot be expressed as PHORS.
%\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}
A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.

>>>>

Many of the recent advances in the development of AST verification methods \cite{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \cite[\S 4.1]{DBLP:journals/toplas/LagoG19}.
(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
By contrast, many probabilistic programming languages allow for richer recursive structures \cite{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.

@@

There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
%Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
%cannot be expressed as PHORS.
%\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}
A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.

### 2021-01-10 LICS21/antitone.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages. (See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.

@@

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)
\, 10
\]
\end{example}

\begin{example}[Escaping spline]
\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . \,
0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
\, 10
\]
\end{example}

@@

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\ F\ \underline x\ \underline y$ is antitone rankable.
\end{example}

>>>>

Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages. (See e.g.~Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.)
Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.

@@

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\cite[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)
\, 10
\]
\end{example}

\begin{example}[Escaping spline]
\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . \,
0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
\, 10
\]
\end{example}

@@

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\ F\ \underline x\ \underline y$ is antitone rankable.
\end{example}

### 2021-01-10 LICS21/applications.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. In \citep{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems.\citep{kiselyov2016problems} Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. Selecting an appropriate addressing scheme can also be important in the context of variational inference.\citep[\S 6.2.2]{paige2016automatic}

>>>>

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems.\cite{kiselyov2016problems} Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. Selecting an appropriate addressing scheme can also be important in the context of variational inference.\cite[\S 6.2.2]{paige2016automatic}

### 2021-01-10 LICS21/confluent-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$. Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first. Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \citep{Bar84,Takahashi1995}. This induction is in reverse, with $k$ decreasing from $n$ to $0$. If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$. If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired. In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.

>>>>

  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$. Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first. Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \cite{Bar84,Takahashi1995}. This induction is in reverse, with $k$ decreasing from $n$ to $0$. If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$. If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired. In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.

### 2021-01-10 LICS21/intro.tex
minor
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. One possible approach to prove almost sure termination is to find some variant on the program state that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. In other words, the program's behavior is used to define an associated ranking supermartingale. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

@@

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk. This restriction can be removed by generalising ranking functions to antitone ranking functions, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly. Our progress condition is equivalent to that given in \citep{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same sub-term multiple times can yield different results. However, with a restricted set of reduction strategies, Church-Rosserness may be regained. A novel addressing scheme for the possible random choices in a program's execution is introduced, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination. This confluent trace semantics has other possible applications as well, for example in inference algorithms.

>>>>

Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. One possible approach to prove almost sure termination is to find some variant on the program state that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. In other words, the program's behavior is used to define an associated ranking supermartingale. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

@@

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk. This restriction can be removed by generalising ranking functions to antitone ranking functions, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly. Our progress condition is equivalent to that given in \cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same sub-term multiple times can yield different results. However, with a restricted set of reduction strategies, Church-Rosserness may be regained. A novel addressing scheme for the possible random choices in a program's execution is introduced, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination. This confluent trace semantics has other possible applications as well, for example in Bayesian inference algorithms.

### 2021-01-10 LICS21/supermartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken by 
%\citep{DBLP:journals/pacmpl/McIverMKK18} 
\cite{DBLP:conf/popl/FioritiH15} and others for imperative programs.

@@

\begin{remark}
Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
\end{remark}

@@

\begin{restatable}{lemma}{rankPast}
%\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
%\end{lemma}
\end{restatable}

\begin{remark}
\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
\end{remark}

>>>>

One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken by 
%\cite{DBLP:journals/pacmpl/McIverMKK18} 
\cite{DBLP:conf/popl/FioritiH15} and others for imperative programs.

@@

\begin{remark}
Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
\end{remark}

@@

\begin{restatable}{lemma}{rankPast}
%\begin{lemma}%[{\cite[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \cite[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
%\end{lemma}
\end{restatable}

\begin{remark}
\Cref{lem:rank-PAST} is a slight extension of \cite[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
\end{remark}
